Hydras \& Co.

### 2022-02-10 doc/gaia-chapter.tex
revision of chapter 10 (continued)

Most used constants and functions will also get standardized
non-ambiguous names.

@@

@@

@@

The functions \texttt{iota} and \texttt{pi} allow us to define
a notion of refinement for constants, functions, predicates and relations.

@@

@@

\subsubsection{Binary operations}
The binary operations $+$ and $\times$ are defined the same way in both libraries. Please note that they require comparison of ordinal terms. So, the proof of the following lemmas applies
compatibility lemmas like \texttt{compare\_ref} (see Section~\vref{sect:gaia-compare-ref}).

\subsection{Order and comparison}
Leibniz equality and strict order on both libraries are
compatible:
\inputsnippets{T1Bridge/eqRef}

@@

\inputsnippets{T1Bridge/ltRef}

@@

@@

\subsection{Porting a theorem from \gaia}
The \texttt{Gaia} library already contains a proof that the multiplication of ordinals less than $\epsilon_0$ is associative.
\emph{From~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}}

@@

From now on, names like  \texttt{T1}, \texttt{omega}, etc. are  bound to \HydrasLib' meaning.

@@

\subsection{Porting theorems from \HydrasLib}

@@

\inputsnippets{T1Bridge/T1leIff}
\inputsnippets{T1Bridge/rewritingRules}
\inputsnippets{T1Bridge/limitbRef}

Please note that in this module, we favour \gaia's notations over hydra's. A snippet like \texttt{alpha < succ beta} should be interpreted as \texttt{T1lt alpha beta}, where \texttt{T1lt} is defined in ~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}.

\texttt{gaia.ssete9.v}. The corresponding statement is
written \texttt{hlt (g2h alpha) (g2h beta)}.

@@

\section{Gaia's proof of well-foundedness}
The library \HydrasLib already contains two proofs
of well-foundedness of the strict order on ordinal terms in
Cantor normal form: a direct proof~\vref{sec:T1Wf} and a proof based on the recursive path ordering~\vref{remark:a3pat}, by
\'Evelyne Contejean. We found it interesting to comment Gaia's proof, which uses tools already present in~\cite{CantorContrib}.

This proof is quite dense and qualified as ``a bit tricky''. We just present here its main steps with the help of the \alectr tool.

The reader is invited to replay the full proof in  \href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}, and to compare with
the proof in Section~\ref{sec:strongly-accessible}.

\begin{remark}
  The following snippets have been generated, not from
  \href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}, but on a provisional copy:
\url{../theories/gaia/nfgaia.v} we made in order to insert comments for \alectr.  This copy is planned to be removed when we are able to add \alectr directives to a \coq source we are not allowed to modify.
\end{remark}

\subsection{Restricted recursion}
As remarked in~\vref{sec:T1Wf}, the order \texttt{T1lt} on \texttt{T1} is not well-founded, because of terms not in normal form.
But the \emph{restriction} of \texttt{T1lt} to terms in normal form \emph{is} well-founded. The following section introduces a  vocabulary in order to reason about statements of the form  ``the restriction of the relation $R$ to a subset $P$ of $A$ is well founded''.

\inputsnippets{nfwfgaia/restrictedRecursion}

The following induction principle is just a variant of \texttt{well\_founded\_induction\_type} for relations defined with \texttt{restrict}.

\inputsnippets{nfwfgaia/restrictedRecursiona}

With $P=\texttt{T1nf}$, this principle will allow us to write proofs by transfinite induction.

\subsection{Main proof steps}

\emph{Notes: the abbreviation \texttt{LT}, and the \texttt{fold LT} and \texttt{unfold LT} tactics,are not in \texttt{gaia}, but  have been added to make the goal display more readable.
The infix notation \texttt{`` \_ < \_''}  is bound to the order \texttt{T1lt} in \gaia, while,  in \HydrasLib,  it is bound to its restriction \texttt{LT} to terms in normal form. }

\inputsnippets{nfwfgaia/nfWfProofa}
We start with a classic structural induction on the type \texttt{T1}.

\inputsnippets{nfwfgaia/nfWfProofaa}
The following \texttt{ssreflect} tactic  pushes a new hypothesis \texttt{Hb} which --- in the same terms as in Section~\vref{sec:strongly-accessible} --- tells that any ordinal less than $a$ is ``strongly accessible''.

\inputsnippets{nfwfgaia/nfWfProofb}

The following sentence gives us the possibility of induct over \texttt{a}'s accessibility.

\inputsnippets{nfwfgaia/nfWfProofbb}

By the following two forward steps, we show that $b$ is accessible, and will be able to prove properties by induction on $b$'s accessibility.

\inputsnippets{nfwfgaia/haveHc}

\inputsnippets{nfwfgaia/haveHd}

\inputsnippets{nfwfgaia/hdProved}

Now, let us consider any ordinal term which is (by \texttt{LT}) less than \texttt{(cons a n b)}.

There are three cases to consider.
\begin{enumerate}
\item It is of the form \texttt{(cons a' n' b')} where \texttt{a'} is strictly less than \texttt{a}.
  \inputsnippets{nfwfgaia/nfWfProofd}

  Then, it is accessible (by hypothesis \texttt{Hb}).

  \item The considered term  is equal to
    \texttt{(cons a n b'')}, where $b''<\omega^a$. 
    \inputsnippets{nfwfgaia/nfWfProofg}

    Then we can replace \texttt{a''} with \texttt{a}, \texttt{n''} with \texttt{n}, and apply the hypotheses \texttt{Hd} and \texttt{He}.

\item
  The term is equal to
  \texttt{(cons a n'' b'')}, where $n''<n$ and $b''<\omega^a$.

  \inputsnippets{nfwfgaia/nfWfProofe}

  By \texttt{Hd} and \texttt{He}, and a few technical lemmas about normal forms,   the term  \texttt{(cons a n b'')} is accessible, and so is \texttt{(cons a n'' b'')}.

\end{enumerate}

In every case, the considered ordinal is accessible. By the definition of accessibility, the ordinal \texttt{(cons a n b)} is accessible, which ends the proof.

>>>>

\subsubsection{Renamings}

Most used constants and functions are also unambiguously renamed.

@@

Please note that in this module, we favour \gaia's notations over hydra's. A snippet like \texttt{alpha < succ beta} should be interpreted as \texttt{T1lt alpha beta}, where \texttt{T1lt} is defined in ~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{\texttt{gaia.ssete9.v}}.

\subsubsection{Translation functions and data refinement}

@@

The following cancel lemmas will be often used in order to simplify sub-terms of the form \texttt{g2h (h2g {\it t})} and \texttt{h2g (g2h {\it t})} which appear in many proofs by rewriting.

@@

\inputsnippets{T1Bridge/h2gEqIff, T1Bridge/g2hEqIff}

Functions \texttt{h2g} and \texttt{g2h} allow us to define
a notion of ``data-refinement''  for constants, functions, predicates and relations. The following definitions express that some
constant, function, relation defined in \HydrasLib ``implements'' the same concept of \gaia.

@@

%%% ICI

@@

\subsubsection{Order and comparison}
Leibniz equality and strict order on both libraries are
compatible:

\inputsnippets{T1Bridge/eqRef}

@@

\inputsnippets{T1Bridge/ltRef, T1Bridge/leRef}

@@

\subsubsection{Cantor normal form}

@@

\subsubsection{Addition and multiplication}

The binary operations $+$ and $\times$ are defined the same way in both libraries. Please note that they require comparison of ordinal terms. So, the proof of the following lemmas applies
compatibility lemmas like \texttt{compare\_ref} (see Section~\vref{sect:gaia-compare-ref}).

\subsection{Importing a theorem from \gaia}
The \texttt{Gaia} library already contains a proof that the multiplication of ordinals less than $\epsilon_0$ is associative.
\emph{From~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}}

@@

In the rest of this module, names like  \texttt{T1}, \texttt{omega}, etc. are  bound to \HydrasLib' meaning.

@@

\subsection{Importing theorems from \HydrasLib}

@@

\inputsnippets{T1Bridge/T1leIff}
\inputsnippets{T1Bridge/rewritingRules}
\inputsnippets{T1Bridge/T1ltIff}

\inputsnippets{T1Bridge/limitbRef, T1Bridge/isSuccRef}

@@

% \section{Gaia's proof of well-foundedness}
% The library \HydrasLib already contains two proofs
% of well-foundedness of the strict order on ordinal terms in
% Cantor normal form: a direct proof~\vref{sec:T1Wf} and a proof based on the recursive path ordering~\vref{remark:a3pat}, by
% \'Evelyne Contejean. We found it interesting to comment Gaia's proof, which uses tools already present in~\cite{CantorContrib}.

% This proof is quite dense and qualified as ``a bit tricky''. We just present here its main steps with the help of the \alectr tool.

% The reader is invited to replay the full proof in  \href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}, and to compare with
% the proof in Section~\ref{sec:strongly-accessible}.

% \begin{remark}
%   The following snippets have been generated, not from
%   \href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}, but on a provisional copy:
% \url{../theories/gaia/nfgaia.v} we made in order to insert comments for \alectr.  This copy is planned to be removed when we are able to add \alectr directives to a \coq source we are not allowed to modify.
% \end{remark}

% \subsection{Restricted recursion}
% As remarked in~\vref{sec:T1Wf}, the order \texttt{T1lt} on \texttt{T1} is not well-founded, because of terms not in normal form.
% But the \emph{restriction} of \texttt{T1lt} to terms in normal form \emph{is} well-founded. The following section introduces a  vocabulary in order to reason about statements of the form  ``the restriction of the relation $R$ to a subset $P$ of $A$ is well founded''.

% \inputsnippets{nfwfgaia/restrictedRecursion}

% The following induction principle is just a variant of \texttt{well\_founded\_induction\_type} for relations defined with \texttt{restrict}.

% \inputsnippets{nfwfgaia/restrictedRecursiona}

% With $P=\texttt{T1nf}$, this principle will allow us to write proofs by transfinite induction.

% \subsection{Main proof steps}

% \emph{Notes: the abbreviation \texttt{LT}, and the \texttt{fold LT} and \texttt{unfold LT} tactics,are not in \texttt{gaia}, but  have been added to make the goal display more readable.
% The infix notation \texttt{`` \_ < \_''}  is bound to the order \texttt{T1lt} in \gaia, while,  in \HydrasLib,  it is bound to its restriction \texttt{LT} to terms in normal form. }

% \inputsnippets{nfwfgaia/nfWfProofa}
% We start with a classic structural induction on the type \texttt{T1}.

% \inputsnippets{nfwfgaia/nfWfProofaa}
% The following \texttt{ssreflect} tactic  pushes a new hypothesis \texttt{Hb} which --- in the same terms as in Section~\vref{sec:strongly-accessible} --- tells that any ordinal less than $a$ is ``strongly accessible''.

% \inputsnippets{nfwfgaia/nfWfProofb}

% The following sentence gives us the possibility of induct over \texttt{a}'s accessibility.

% \inputsnippets{nfwfgaia/nfWfProofbb}

% By the following two forward steps, we show that $b$ is accessible, and will be able to prove properties by induction on $b$'s accessibility.

% \inputsnippets{nfwfgaia/haveHc}

% \inputsnippets{nfwfgaia/haveHd}

% \inputsnippets{nfwfgaia/hdProved}

% Now, let us consider any ordinal term which is (by \texttt{LT}) less than \texttt{(cons a n b)}.

% There are three cases to consider.
% \begin{enumerate}
% \item It is of the form \texttt{(cons a' n' b')} where \texttt{a'} is strictly less than \texttt{a}.
%   \inputsnippets{nfwfgaia/nfWfProofd}

%   Then, it is accessible (by hypothesis \texttt{Hb}).

%   \item The considered term  is equal to
%     \texttt{(cons a n b'')}, where $b''<\omega^a$. 
%     \inputsnippets{nfwfgaia/nfWfProofg}

%     Then we can replace \texttt{a''} with \texttt{a}, \texttt{n''} with \texttt{n}, and apply the hypotheses \texttt{Hd} and \texttt{He}.

% \item
%   The term is equal to
%   \texttt{(cons a n'' b'')}, where $n''<n$ and $b''<\omega^a$.

%   \inputsnippets{nfwfgaia/nfWfProofe}

%   By \texttt{Hd} and \texttt{He}, and a few technical lemmas about normal forms,   the term  \texttt{(cons a n b'')} is accessible, and so is \texttt{(cons a n'' b'')}.

% \end{enumerate}

% In every case, the considered ordinal is accessible. By the definition of accessibility, the ordinal \texttt{(cons a n b)} is accessible, which ends the proof.
